The role of advection versus diffusion in intraventricular transport

### 2025-03-18 paper/main.tex
more references and intro work

\keywords{cerebrospinal fluid, brain ventricles, transport, finite element methods}

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid mostly produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment, but for which there is 
little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are evolutionarily conserved, microscopic hair-like
structures that beat in a coordinate pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution, Ringers2020Role}.
Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes primarily to flow near the
ventricular surface in the human brain~\cite{Siyahhan2014FlowVentricles}.
On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled}.

@@

%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%

Still an underexplored topic, how the motile cilia contribute to overall CSF flow
and solute transport remains poorly understood. Experimental endeavors are
challenging to pursue due to technical limitations: the microscopic size
and high beating frequencies of the cilia make them difficult to monitor 
in vivo in humans. Studies on animal models such as mice and
zebrafish have provided unequivocal evidence that cilia are crucial
for CSF flow in these animals~\cite{Olstad2019CiliaryDevelopment, Faubel2016Cilia}.
It is clear that cilia affects advective intraventricular transport 
to some extent. However a quantification of this effect, as well as of to which
extent diffusion plays a role in transporting solutes, remains
yet to be performed.

Studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction will alter CSF composition and lead to brain
disorders. The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\texbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}

>>>>

\keywords{cerebrospinal fluid, brain ventricles, cilia, transport, finite element methods}

@@

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. Produced mostly by the choroid
plexus and partly by brain interstitial fluid,
CSF is a clear liquid that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013, Brodal2016}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{Ringers2020Role, fame2020emergence, del2010ependymal}.
Disruption of the CSF circulation is detrimental to brain health and disturbs
cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment~\cite{duy2022rethinking, kahle2024paediatric, wallmeier2022role},
but for which there is little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile,
Ibanez-Tallon2002LossHydrocephalus},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Motile cilia are hair-like structures that line the ventricular surface of the brain,
known as the ependyma~\cite{mitchell2007evolution, Ringers2020Role, spassky2013motile}. 
The cilia beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%
%  Studies on animal models such as mice and
% zebrafish have provided unequivocal evidence that cilia are crucial
% for CSF flow in these animals~\cite{Olstad2019CiliaryDevelopment, Faubel2016Cilia}.
% It is clear that cilia affects advective intraventricular transport 
% to some extent. However a quantification of this effect, as well as of to which
% extent diffusion plays a role in transporting solutes, remains
% yet to be performed.

To date, it remains poorly understood how motile cilia contribute to 
the overall fluid flow and solute transport within the brain.
Experimental endeavors are challenging to pursue due to
technical limitations: the microscopic size and relatively 
high beating frequencies of the cilia~\cite{ocallaghan2012analysis, mahuzier2018ependymal,
Jeong2022MeasurementTelencephalon, Olstad2019CiliaryDevelopment, spassky2013motile, DGama2021Diversity},
as compared to other fluid dynamic processes in the brain that are slower and span larger
spatial scales~\cite{Vinje2019RespiratoryMeasurements, eide2024functional, fultz2019coupled, mestre2018flow},
make cilia difficult to monitor in vivo in humans.
Until now, most studies have analyzed one physical mechanism at a time, especially in mammals.
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles~\cite{Faubel2016Cilia, Sawamoto2006NewBrain, ibanez2003beat};
however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans~\cite{feinberg1987human, battal2011cerebrospinal,
yamada2015current}, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain~\cite{Siyahhan2014FlowVentricles}. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction alters CSF composition and thus leads to brain disorders.

The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays, roth2025structure},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b, c.} Particle tracking data visualizing CSF flow directionality (\textbf{b}) and
    velocity magnitudes in the middle ventricle.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries.
    \textbf{h--j.} Streamlines simulated with the baseline flow model at
    a time a time 25\% (\textbf{h}), 50\% (\textbf{i}) and 75\% (\textbf{j})
    into the cardiac cycle. Times indicated in panel \textbf{g}.
    %t=0.112613 s %t=0.225225 s %t=0.315315 s
    Scale bars 50 \textmu m.
    }
    \label{fig:fig1}
\end{figure}
